Estimating demographic bias on tests of children's early vocabulary

### 2022-01-30 paper/cogsci.Rmd
making figures horizontal

```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs. age by demographic group, from the Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors)
  #theme(legend.position = "top")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "SES", type=ses_colors) #+ 
  #theme(legend.position = "top")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "ethnicity", type=eth_colors) #+ 
  #theme(legend.position = "top")

@@

Figures 2-4 show GLIMMERs for a selection of CDI items for sex, SES, and race, respectively.
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

```{r sex-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model. Words at the top are more well-known by females, while those at the bottom are more known for males."}
#img <- png::readPNG("figs/lab_logo_stanford.png")
#grid::grid.raster(img)
knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
```

```{r ses-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model. Words at the top are more known by children in low-SES families, while those at the bottom are more known by those in high-SES families."}
knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
```

```{r eth-glimmer, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the race bias model. Words at the top are more known by non-white children, while those at the bottom are more known by white children."}
knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
```

@@

```{r pruned-coefs, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.4, fig.height=4.75, set.cap.width=T, num.cols.cap=1, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
load(here("data/pruned_demo_coefs.Rdata"))

pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
  pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
                           names_to=c("type", "demo"), names_sep="_") %>%
  pivot_wider(names_from=type, values_from=value) %>%
  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo)))

@@

ggarrange(pr1, pr2, pr3, ncol=1, common.legend = T)
```

@@

Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
For each demographic factor, we pruned items with difficulty difference from the mean at varying thresholds, from >.25 SD to >3 SD, in increments of .25 SD. 
At each SD threshold, a potentially different subset of items are excluded for each model, but with the goal of creating a single CDI that is less biased on all dimensions, we pruned the union of the subsets excluded from each model.
For example, pruning >2SD from the mean of each model excluded 25 sex-biased items, 39 ME-biased items, and 31 race-biased items, with their union being 76 unique items.
Figure 5 shows the size of each demographic effect at different exclusion thresholds.

@@

```{r demo-ability, include=T, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}

>>>>

```{r baseline-ability-vs-age-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Language ability vs. age by demographic group, from the baseline Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors) +
  theme(legend.position = "top")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "SES", type=ses_colors) + 
  theme(legend.position = "top")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "ethnicity", type=eth_colors) + 
  theme(legend.position = "top")

ggarrange(p1, p2, p3, ncol=3, nrow=1)
```

@@

Figure 2 shows GLIMMERs for a selection of CDI items for sex (left), maternal education (middle), and race (right).
The full GLIMMERs, with all 680 CDI:WS items, are available on [OSF](https://osf.io/57rsw/?view_only=2b6ecb61fe08458293af7421d276932a), but were too large to include here.
It is important, however, to inspect the full plots, for if there is a cluster of items in a GLIMMER, the analyst may conclude that these items are strong candidates for DIF on that dimension.
For example, there is some clustering at the top and bottom of Figure 2: at the top, "vagina", "tights", "dress (object)", and "doll" form a cluster of items that are much more well-known for females, while at the bottom, "penis" stands out as much more well-known by males.
For the maternal education and race GLIMMERs (Figures 3 and 4)--and in the rest of the full sex GLIMMER, there are not clusters, but rather a continuum of smoothly varying differences with overlap. 
This makes identifying items with DIF quite difficult, as different methods are likely to yield inconsistent results [@stenhaug2021treading].
Hence, we next characterized the distributions of the item-level group difficulty differences, and measured the influence of pruning a varying number of items on the demographic effect sizes.

```{r glimmer-combo, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.width=6.5, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model. Words at the top are more well-known by females, while those at the bottom are more known for males."}
knitr::include_graphics("figs/smGLIMMER_combo.pdf")
```

```{r sex-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.2, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the sex bias model. Words at the top are more well-known by females, while those at the bottom are more known for males."}
knitr::include_graphics("figs/smGLIMMER_sex_prodWS.pdf")
```

```{r ses-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the SES bias model. Words at the top are more known by children in low-SES families, while those at the bottom are more known by those in high-SES families."}
knitr::include_graphics("figs/smGLIMMER_ses_prodWS.pdf")
```

```{r eth-glimmer, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3, fig.height=3.5, set.cap.width=T, num.cols.cap=1, fig.cap = "GLIMMER plot of a sample of CDI:WS words from the race bias model. Words at the top are more known by non-white children, while those at the bottom are more known by white children."}
knitr::include_graphics("figs/smGLIMMER_eth_prodWS.pdf")
```

@@

```{r pruned-coefs-hor, fig.env = "figure*", out.width="\\linewidth", fig.pos = "h", fig.align='center', fig.width=6.5, fig.height=2.8, set.cap.width=T, num.cols.cap=2, fig.cap = "Size of demographic effects (regression coefficients) with different pruning thresholds."}
load(here("data/pruned_demo_coefs.Rdata"))
# horizontal version
pdat <- prune_dat %>% rename(c(beta_sex=sex_beta, beta_ses=ses_beta, beta_eth=eth_beta)) %>%
  pivot_longer(c("extreme_sex","extreme_ses","extreme_eth","beta_sex","beta_ses","beta_eth"), 
                           names_to=c("type", "demo"), names_sep="_") %>%
  pivot_wider(names_from=type, values_from=value) %>%
  mutate(Dimension = ifelse(demo=="ses", "maternal ed", ifelse(demo=="eth", "race", demo)))

@@

ggarrange(pr1, pr2, pr3, nrow=1, ncol=3, common.legend = T)
```

@@

Given the smoothly-varying distributions of DIF shown by each demographic factor, we chose to evaluate multiple thresholds for pruning the extreme-valued items from each distribution.
For each demographic factor, we pruned items with difficulty difference from the mean at varying thresholds, from >.25 SD to >3 SD, in increments of .25 SD. 
At each SD threshold, a potentially different subset of items are excluded for each model, but with the goal of creating a single CDI that is less biased on all dimensions, we pruned the union of the subsets excluded from each model.
For example, pruning >2SD from the mean of each model excluded 25 sex-biased items, 39 ME-biased items, and 31 race-biased items, with their union being 76 unique items.
Figure 3 shows the demographic effects ($\beta$s) at different exclusion thresholds.

@@

```{r demo-ability, include=F, out.width="\\linewidth", fig.env = "figure", fig.pos = "h", fig.align='center', fig.width=3.5, fig.height=7.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Scatterplots showing item difficulty per demographic group along each dimension."}
